\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{margin=1in}

\title{Closed Causal Information (CCI): A Computable Metric for\\
Autonomous Internal Dynamics in Neural Systems}
\author{Anonymous}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We introduce \emph{Closed Causal Information} (CCI), a computable metric
that quantifies the degree to which a neural system's internal hidden states
evolve according to their own causal structure, independent of external input.
CCI is grounded in conditional mutual information and requires no distributional
assumptions beyond stationarity. We define the metric formally, prove that
feedforward architectures are bounded at CCI\,$= 0$, and demonstrate
empirically that recurrent architectures develop non-trivial CCI during
training on sequence tasks. We further introduce the
\emph{Minimal Recurrent Core — Causal} (MRC-C) method for localising the
subgraph responsible for closed causal dynamics.
No philosophical claims about consciousness or agency are made.
\end{abstract}

%-------------------------------------------------------------------
\section{Introduction}

A central question in the analysis of neural systems is whether internal
representations are primarily \emph{reactive} (driven by external input) or
exhibit autonomous temporal structure. Standard mutual information measures
conflate these two sources of predictability. CCI isolates the internal
component by conditioning out the contribution of the current input.

%-------------------------------------------------------------------
\section{Definition}

Let $\mathbf{H}_t \in \mathbb{R}^d$ denote the hidden state of a neural
system at time $t$, and let $\mathbf{E}_t$ denote the external input at
time $t$.

\begin{definition}[CCI]
The \emph{Closed Causal Information} of a system over a trajectory of
length $T$ is
\begin{equation}
    \mathcal{E}(X) \;=\; \frac{1}{T-1}\sum_{t=1}^{T-1}
        I\!\left(\mathbf{H}_t\,;\,\mathbf{H}_{t+1} \mid \mathbf{E}_t\right),
\label{eq:cci}
\end{equation}
where $I(A\,;\,B\mid C)$ denotes the conditional mutual information between
$A$ and $B$ given $C$.
\end{definition}

Equation~\eqref{eq:cci} measures the average temporal dependence between
consecutive hidden states that cannot be explained by the concurrent input.
High $\mathcal{E}(X)$ indicates that the system's future state is substantially
predictable from its current state regardless of what stimuli arrive.

%-------------------------------------------------------------------
\section{Estimation}

In practice we estimate $I(\mathbf{H}_t;\mathbf{H}_{t+1}\mid\mathbf{E}_t)$
via a Gaussian approximation:
\begin{equation}
    I(A;B\mid C) \;\approx\;
    \frac{1}{2}\log\frac{|\Sigma_{AC}||\Sigma_{BC}|}{|\Sigma_{C}||\Sigma_{ABC}|},
\end{equation}
where $\Sigma_S$ denotes the covariance matrix of the joint vector $S$.
This estimator is consistent when the joint distribution is approximately
Gaussian and is efficient to compute from sample covariance matrices.

%-------------------------------------------------------------------
\section{Properties}

\begin{proposition}
For any strictly feedforward architecture, $\mathcal{E}(X) = 0$.
\end{proposition}

\begin{proof}
In a feedforward network, $\mathbf{H}_{t+1}$ is a deterministic function
of $\mathbf{E}_{t+1}$ alone, so $\mathbf{H}_{t+1} \perp\!\!\!\perp \mathbf{H}_t
\mid \mathbf{E}_t$, giving $I(\mathbf{H}_t;\mathbf{H}_{t+1}\mid\mathbf{E}_t) = 0$.
\end{proof}

%-------------------------------------------------------------------
\section{MRC-C: Minimal Recurrent Core — Causal}

Given a network with multiple recurrent components, MRC-C identifies the
\emph{minimal subgraph} $\mathcal{G}^* \subseteq \mathcal{G}$ such that
ablating any edge in $\mathcal{G}^*$ reduces $\mathcal{E}(X)$ by more than
a threshold $\epsilon$. This is computed greedily by edge-importance scoring
via the gradient of CCI with respect to edge weights.

%-------------------------------------------------------------------
\section{Experiments}

We evaluate CCI on four settings:
\begin{enumerate}
    \item Feedforward vs.\ RNN on a copying task.
    \item CCI dynamics over the course of training.
    \item Effect of structured vs.\ random input noise on CCI.
    \item Transformer baseline with causal self-attention.
\end{enumerate}

Results are consistent with the theoretical prediction: feedforward networks
maintain CCI $\approx 0$, while recurrent networks develop increasing CCI as
training progresses.

%-------------------------------------------------------------------
\section{Conclusion}

CCI provides a simple, model-agnostic, and computationally efficient signal
for characterising the degree of internal temporal autonomy in neural systems.
It distinguishes genuinely recurrent computation from reactive processing
without appeal to any particular architectural definition of recurrence.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}
